Covariate analysis of MSSM reprocessed counts (with TMM and SVA normalisation)

### 2017-01-23 code/Rmd/MSSM_RNASeq_Reprocessed_TMM_SVA.Rmd
Added markdown to process reprocessed RNASeq from MSBB with TMM and SVA normalization

---
title: "Covariate analysis of MSSM reprocessed counts (with TMM normalisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)

@@

knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_Reprocessed_TMM.Rmd",
                                 parentId = "syn5570248",
                                 entityName = 'MSBB Reprocessed RNASeq TMM')
```

@@

thisFileName <- 'MSSM_RNASeq_Reprocessed_TMM.Rmd'

@@

# Find significant covariates
adjustCovars = preAdjustedSigCovars$significantCovars
```
Significant covariates to adjust at FDR 0.1 are `r paste(gsub('_','\\\\_',adjustCovars), collapse= ',')`
```{r preAdjustedSigCovars.NULL.ALL, fig.width=25, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
```{r iterative.adjusted.voom.normalisation, results='asis'}
postAdjustCovars = c('BrodmannArea.CDR', 'batch', 'SEX')
randomCovariates = 'individualIdentifier'

# Assign residual covariates
residualSigCovars = preAdjustedSigCovars
significantCovars = residualSigCovars$significantCovars
covariatesEffects = residualSigCovars$Effects.significantCovars
covariatesEffects = covariatesEffects[setdiff(significantCovars, c(postAdjustCovars, randomCovariates))]

postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))

loopCount = 0 
while(length(significantCovars)!=0 && loopCount <= 100){
  writeLines(paste('Using following covariates in the model:', 
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects and individualIdentifier as random effects'))

@@

@@

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, randomCovariates))

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES,
                                                'all adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  significantCovars = residualSigCovars$significantCovars
  covariatesEffects = residualSigCovars$Effects.significantCovars
  covariatesEffects = covariatesEffects[setdiff(significantCovars,c(postAdjustCovars, randomCovariates))]

  postAdjustCovars = c(postAdjustCovars,names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}
modelStr <-paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=','), 'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr, 'and Donor_ID as random effects')
```
`r tmp`
### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
Adding "BrodmannArea.CDR" back to the residuals
```{r addVarsBack}
varsToAddBackIn <-  grep("BrainRegion.Diagnosis", colnames(FIT$coefficients), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(FIT$coefficients[,varsToAddBackIn]))
```

@@

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'MSBB Reprocessed RNASeq TMM', parentId = parentId)
CODE <- synStore(CODE)

>>>>

---
title: "Covariate analysis of MSSM reprocessed counts (with TMM and SVA normalisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)

@@

knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_Reprocessed_TMM_SVA.Rmd",
                                 parentId = "syn5570248",
                                 entityName = 'MSBB Reprocessed RNASeq TMM SVA')
```

@@

thisFileName <- 'MSSM_RNASeq_Reprocessed_TMM_SVA.Rmd'

@@

# Find significant covariates
adjustCovars = preAdjustedSigCovars$significantCovars
```
Significant covariates to adjust at FDR 0.1 are `r paste(gsub('_','\\\\_',adjustCovars), collapse= ',')`
```{r preAdjustedSigCovars.NULL.ALL, fig.width=25, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (custom design)
Since many covariates are correlated, re-normalising COVARIATES with a custom design matrix (see syn8049649 for details)
```{r iterative.adjusted.voom.normalisation, results='asis'}
postAdjustCovars = c('BrodmannArea.CDR', 'batch', 'SEX', 'RIN', 'rRNA.rate', 'PMI', 'AOD')
randomCovariates = 'individualIdentifier'

writeLines(paste('Using following covariates in the model:', 
                 paste(postAdjustCovars, collapse=', '),
                 'as fixed effects and individualIdentifier as random effects'))

@@

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
```
`r tmp`
### SVA 
Estimate surrogate variables using SVA package
```{r sva}
MODEL0 = DM1$design[,-grep('BrodmannArea.CDR',colnames(DM1$design))] # Get null model by removing variable of interest
MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
MODEL1 = DM1$design
MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]

## Calculate the number of surrogate variables
NUM.SV = sva::num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1,
                     method = 'be', seed = 123456, B = 30)

SURR.VAR = sva::sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0,
                    n.sv = NUM.SV, B = 30)$sv

colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
rownames(SURR.VAR) = rownames(DM1$design)

DM1$design = cbind(MODEL1, SURR.VAR)

# Calculate correlation values of random effects
correlation = parallelDuplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION, block = COVARIATES$Donor_ID)

# Re-estimate voom weights
VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F, 
                                     block = COVARIATES$Donor_ID, correlation = correlation$cor)

@@

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              COVARIATES,
                                              'all adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```
### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
Adding "BrodmannArea.CDR" back to the residuals
```{r addVarsBack}
varsToAddBackIn <-  grep("BrodmannArea.CDR", colnames(FIT$coefficients), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(FIT$coefficients[,varsToAddBackIn]))
```

@@

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'MSBB Reprocessed RNASeq TMM SVA', parentId = parentId)
CODE <- synStore(CODE)
